The `forestecology` R package for modeling interspecies competition between trees

### 2021-02-15 etc/paper/forestecology.Rmd
Done "add spatial information" section

# Set ggplot defaults output:
if (!is_html_output()) {
  # Grey theme:
  ggplot2::theme_set(theme_light())
  # Color scheme
  scale_colour_discrete <- scale_colour_viridis_d
}

@@

```{r load-packages}
library(tidyverse)
library(lubridate)
library(sf)
library(forestecology)
```

@@

We now encode spatial information to the `growth_df` data frames. First, in order to control for study region edge effects, we add "buffers" to the periphery of the study region (cite Waller?). Our model of interspecific competition relies on a spatial definition of who the competitor trees are for focal trees of interest. Since certain explanatory variables such as basal area are cumulative, we must ensure that all trees being modeled are not biased to have different neighbor structures. This is a particular concern for trees at the boundary of study regions, which will not have the same number of neighbors as trees in the internal part of the study region.

Second, our ultimate method for model assessment will rely on estimates of model error as generated by cross-validation. Conventional cross-validation schemes assign observations to folds by resampling individual observations at random. However, underlying this scheme is an assumption that the observations are independent. In the case of forest census data, observations exhibit spatial autocorrelation, and thus this dependence must be incorporated in our resampling scheme in spatial cross-validation @roberts_cross-validation_2017 @pohjankukka_estimating_2017 We will therefore associate portions of the study region to spatial folds.

To these two ends, we define two constants, both of which are in the same units as the `gx` and `gy` variables (most often meters).

```{r}
comp_dist <- 7.5
cv_fold_size <- 100
```

The first constant is `comp_dist` which defines the maximum distance for a tree's competitive neighborhood. Trees within this distance of each other are assumed to compete while those farther than this distance apart do not. Put differently, all trees within `comp_dist` of a focal tree will be considered its competitors (see below). Other studies have estimated the value of `comp_dist`; we use an average of estimated values @canham_neighborhood_2004, @uriarte_spatially_2004, @tatsumi2013, @canham_neighborhood_2006.

Furthermore, `comp_dist` will define the size of all buffers considered, which will be encoded as a binary variable `buffer` as computed by the `add_buffer_variable()` function. This function takes as input the main `growth_df` data frame, the `size` of the buffer which we set as `comp_dist`, and the boundary of the study region encoded as a simple features polygon @pebesma_simple_2018. DESCRIBE SF PACKAGE. In the Big Woods example below we will use a pre-loaded simple features polygon while for the SCBI example we present example code on how to manually construct one.

The second constant is `cv_fold_size` which defines the length and width of the spatial folds (note that for now the spatial folds are restricted be squares). We will then use this constant to associate each observed tree to one of $k$ folds in the respective study region. In the Big Woods example below we will use the `blockCV` R package that has implemented spatial cross-validation while for the SCBI we will do this manually @valavi_blockcv_2019

First, we indicate which trees are part of the buffer. This necessitates information about the study region boundary. In this case, we use a `sf_polygon` object `study_region_bw` which comes pre-loaded in the `forestecology` packages. After loading `study_region_bw`, we illustrate the results of the `add_buffer_variable()` function in Figure \ref{fig:bw-define-buffer}. Trees on the periphery denote with lighter colors are part of the buffer and will not be considered as "focal" trees of interest going forward; they will only be considered as competitor trees.

```{r bw-define-buffer, out.width="100%", fig.cap="Buffer region for Big Woods study region."}
data(study_region_bw)

growth_bw <- growth_bw %>%
  add_buffer_variable(direction = "in", size = comp_dist, region = study_region_bw)

ggplot() +
  geom_sf(data = growth_bw %>% sample_frac(0.2), aes(col = buffer), size = 0.5)
```

Second, we associate each tree to spatial cross validation folds. In this case, we use the `spatialBlock()` function from the `blockCV` package to define the spatial grid which

THIS IS A MESS. We use the  @valavi_blockcv_2019, whose elements will act as the folds in our leave-one-out (by "one" we mean "one grid block") cross-validation scheme. The upshot here is we add `foldID` to `growth_df` which identifies which fold each individual is in, and the creation of a `cv_grid_sf` object which gives the geometry of the cross validation grid.

```{r bw-define-cv-folds, out.width="100%", fig.cap="blockCV function output."}
set.seed(76)
bw_spatialBlock <- spatialBlock(
  speciesData = growth_bw, theRange = cv_fold_size, k = 28, xOffset = 0.5, 
  yOffset = 0, verbose = FALSE, showBlocks = FALSE
)
```

Then add foldID to each tree

```{r}
growth_bw <- growth_bw %>%
  mutate(foldID = bw_spatialBlock$foldID)
```

```{r bw-define-cv-folds-with-trees, out.width="100%", fig.cap="Inspect blocks closely."}
# Visualize grid. Why does fold 19 repeat?
ggplot() +
  geom_sf(data = bw_spatialBlock$blocks %>% st_as_sf()) + 
  geom_sf(data = growth_bw %>% sample_frac(0.2), 
          aes(col = factor(foldID)), size = 0.1, show.legend = FALSE) + 
  geom_sf_text(data = bw_spatialBlock$blocks %>% st_as_sf(), 
               aes(label = folds))
```

Then remove empty folds

```{r}
growth_bw <- growth_bw %>%
  filter(!foldID %in% c(19, 23, 21, 17, 8, 19)) %>%
  mutate(foldID = factor(foldID))
```

Separately, we save the spatial cross-validation grid as an `sf_polygon` object `blocks_bw`

```{r}
blocks_bw <- bw_spatialBlock$blocks %>%
  st_as_sf()
```

First, we indicate which trees are part of the buffer. In this case however we manually define the study region boundary based on the subregion we defined in Section \ref{scbi-data} and create an `sf_polygon` object using the `sf_polygon()` function from the `sfheaders` package. Figure \ref{fig:scbi-define-buffer} displays the resulting buffer trees.

```{r scbi-define-buffer, out.width="50%", fig.cap="Buffer region for SCBI study region."}
study_region_scbi <- tibble(
  x = c(0, 300, 300, 0, 0),
  y = c(300, 300, 600, 600, 300)
) %>%
  sf_polygon()

growth_scbi <- growth_scbi %>%
  add_buffer_variable(direction = "in", size = comp_dist, region = study_region_scbi)

ggplot() +
  geom_sf(data = growth_scbi, aes(col = buffer), size = 0.5)
```

Second, we associate each tree to spatial cross validation folds. In this case we manually define a spatial crossvaliation grid. Figure \ref{fig:scbi-define-cv-folds} displays the resulting cross-validation folds along with the buffer from Figure \ref{fig:scbi-define-buffer}.

Here we manually define the spatial cross-validation grid as an `sf_polygon` object `scbi_cv_grid`

```{r}
fold1 <- rbind(c(0, 300), c(150, 300), c(150, 600), c(0, 600))
fold2 <- rbind(c(150, 300), c(300, 300), c(300, 600), c(150, 600))

blocks_scbi <- bind_rows(
  sf_polygon(fold1),
  sf_polygon(fold2)
) %>%
  mutate(folds = c(1, 2) %>% factor())
```

```{r scbi-define-cv-folds, out.width="50%", fig.cap="Buffer region for SCBI study region."}
SpatialBlock_scbi <- spatialBlock(
  speciesData = growth_scbi, k = 2, selection = "systematic", blocks = blocks_scbi,
  showBlocks = FALSE, verbose = FALSE
)

# Add foldID to each tree
growth_scbi <- growth_scbi %>%
  mutate(foldID = SpatialBlock_scbi$foldID %>% factor())

ggplot() +
  geom_sf_text(data = growth_scbi, aes(label = foldID, col = buffer)) +
  geom_sf(data = blocks_scbi, fill = "transparent")
```

>>>>

# Set ggplot defaults output:
if (!is_html_output()) {
  # Grey theme:
  ggplot2::theme_set(theme_light())
  # Color scheme
  # scale_colour_discrete <- scale_colour_viridis_d
}

@@

```{r load-packages}
library(tidyverse)
library(lubridate)
library(sf)
library(forestecology)
library(blockCV)
```

@@

The next step in our analysis sequence is to add spatial information to our main `growth_scbi` data frame. The first element of spatial information we add is a "buffer region" to the periphery of the study region. Since some of our model's explanatory variables such as competitor basal area are cumulative, we must ensure that all trees being modeled are not biased to have different neighbor structures. This is of particular concern for trees at the boundary of study regions, which will not have the same number of neighbors that act as competitors as trees in the internal part of the study region. In order to account for such edge effects only trees who are not part of this buffer region, i.e. are part of the interior of the study region, will have their growths modeled @waller_applied_2004.

Our model of interspecific competition relies on a spatial definition of who the competitor trees are for focal trees of interest: all trees within a distance `comp_dist` of a focal tree are considered its competitors (assuming the same units as the `gx` and `gy` location variables). In our case we set this value below at 7.5m, a value informed by @canham_neighborhood_2004 @uriarte_spatially_2004 @canham_neighborhood_2006. Using this value along with a manually constructed `sf` object representation of the study region's boundary, we apply the `add_buffer_variable()` to our `growth_scbi` data frame to add a `buffer` boolean variable: all trees who have `buffer` set to `FALSE` will be our focal trees whose growths are modeled, whereas those with `buffer` set to `TRUE` will only be considered as competitor trees whose growth will not be modeled.

```{r}
# Define buffer region using competitive distance range
comp_dist <- 7.5

study_region_scbi <- tibble(
  x = c(0, 300, 300, 0, 0),
  y = c(300, 300, 600, 600, 300)
) %>%
  sf_polygon()

growth_scbi <- growth_scbi %>%
  add_buffer_variable(size = comp_dist, region = study_region_scbi)
```

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, underlying this algorithm is an assumption that the observations are independent. In the case of forest census data, observations exhibit spatial autocorrelation. This spatial dependence is incorporated into the cross-validation algorithm by randomly resampling spatial blocks of trees @roberts_cross-validation_2017 @pohjankukka_estimating_2017. We therefore associate each observed tree to one of $k$ spatial folds. In the example below, we first manually define two folds that partition the study region as an `sf` object. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct fold (saved in the `foldID` variable) @valavi_blockcv_2019. \footnote{In the Appendix we present an example where the folds themselves are also created using the `spatialBlock()` function given a specified `cv_block_size`.}

```{r}
# Manually define spatial blocks to act as folds
fold1 <- rbind(c(0, 300), c(150, 300), c(150, 600), c(0, 600))
fold2 <- rbind(c(150, 300), c(300, 300), c(300, 600), c(150, 600))

blocks_scbi <- bind_rows(sf_polygon(fold1), sf_polygon(fold2)) %>%
  mutate(folds = c(1, 2) %>% factor())

# Associate each observation to a fold
SpatialBlock_scbi <- spatialBlock(
  speciesData = growth_scbi, k = 2, selection = "systematic", 
  blocks = blocks_scbi, showBlocks = FALSE, verbose = FALSE
)

growth_scbi <- growth_scbi %>%
  mutate(foldID = SpatialBlock_scbi$foldID %>% factor())
```

Figure \ref{fig:scbi-spatial-information} illustrates the net effect of adding these two elements of information to the `growth_scbi` data frame. The location of each tree is marked with an integer indicating which fold it belongs to, where the folds are marked with solid lines. The color of each digit indicates whether the tree is part of the buffer region (and thus will only be considered as a competitor tree in our model) or is part of the interior of the study region (and thus is a focal tree whose growth is of modeled interest).

```{r scbi-spatial-information, out.width="66%", fig.cap="Buffer region and cross-validation block information for SCBI data."}
ggplot() +
  geom_sf_text(data = growth_scbi, aes(label = foldID, col = buffer), alpha = 0.2) +
  geom_sf(data = blocks_scbi, fill = "transparent")
```

TODO: Do we talk about `add_buffer_variable(direction = "out")` when each fold gets its turn being the training data?
